Efficiency

### 2018-01-21 DraftExec.tex
axial mixed update, and john and jim edits

It is clear that the optimisation of a system for minimising cavitation, and hence improving acoustic performance, is a separate and independent design requirement to that of optimising a system for propulsive efficiency.  These two design requirements can be found either to be in tension, or alignment, in different regimes.  At high speeds, where cavitation tends to have severe impacts on propulsive efficiency of open propellers, and high mass-flow (which leads to good efficiency) can be achieved through compact ducts and pumps, pumpjets offer very good performance on both metrics.  Hence, in this regime the two design requirements tend to align relatively well in favour of pump-jets.  However, in very low-speed, low-load environments, where cavitation effects are negligible, and larger areas need to be engaged achieve good mass-flow and efficiency, the two design requirements align relatively well in favour of an open propeller.  In a variety of regimes (including at quite high speeds), the two requirements find themselves in tension and needing to be balanced \parencite{gearhart1966selection}.

@@

For context, the equations developed by Wislicenus are used for the purposes generally of selecting the ideal propulor diameter, which can be made larger in the case of a propeller to access higher ideal efficiencies.  This is an opportunity which is constrained in the case of jets because the efficiency diminishes at very low values of $\Delta V / V_0$.  This is quite a different constraint to other "practical" considerations as some have suggested, such as weight, or space constraints.  Open propellers may advantageously choose to lower their values of $\Delta V / V_0$ as low as 0.1 in order to obtain ideal efficiencies \parencite{wislicenus1973}.  This still provides a hard limit on potential improvements that can be attained for a pump-jet that has non-zero duct loss.

@@

In order to begin to model the likely impact of propulsion system choices on system peformance characteristics such as dived range and endurance, it is necessary to adopt some plausible curves indicating likely levels of efficiency across the operating speed range, in this case assumed to be up to 20kt.

In this case, three possible curves have been selected for each system, representing a high, low, and central assumption for each, with peak efficiency for the propeller at about 65\% at near 5kt, and jet efficiency around 27\% at the sames speed.  The central assumption has been taken in both cases from the study undertaken by BMT and Rolls Royce, as in \cite{giles2010}, and indicated in \ref{fig:BMTShape.png}, since this represents the most direct comparison made as a simple function of waterspeed, with comparable technologies.  The technologies naturally will not be identical.  It is probable that the pumpjet of the submarine will be a purely axial-flow pump, as opposed to a mixed flow pump, but this difference in turbomachinery is unlikely to result in a overall loss of efficiency \parencite{hamilton1997}.  The intake duct may be somewhat longer, increasing duct losses, but the degree of diffusion in the duct may be lower, reducing the adverse pressure gradient.  It is probable that the the ship propeller would have five blades, where the submarine would likely have seven or nine. However, being  for a warship designed to travel at similar speeds to a nuclear submarine, with relatively unconstrained propeller size, and the pumpjet ejecting into the water, the technologies are likely to be far closer than any other publically available comparison. Overall, however, its considered that having an experimtally verified curve comparing very similar technologies as a direct function of waterspeed provides a far more certain basis for comparison than attempting to derive different curves \textit{ab initio} using theortical dimesnionless curves attempting to estimate how the proposed designs might correspond to them.

Variations from this central line have been made with consideration for some plausible trade-offs, which are intended to be indicative, but plausible.  In the case of the high pumpjet assumption, it is assume that a substantially larger intake area has been chosen in order to achieve higher efficiencies at lower speeds, at the cost of peak ideal efficiency, which might be realised not too far above 20kt for such a design.  The low assumption for the pumpjet is still represents higher efficiencies at equivalent speeds than those given in \cite{fujisawa1995} for a waterjet. It is assumed that for both the central and low pumpjet curves, peak efficiency would not be achieved until a speed considerably above 20kt.  All of the efficiency curves reflect a smaller decline in efficiency than the indicative assumption used by Andrew Davies, namely that efficiency might reduce by a factor of four between transit and patrol speeds, if those speeds also differed by a factor of four \cite{davies2017}.

@@

It is worthwhile considering that alternative types of battery may be on offer for submarines.  In particular, lithium ion batteries were offered by TKMS in their bid for SEA 1000, and are being installed in Type 212 submarines operated by the German navy, as well as contracted for the new Norwegian submarines, as well as Israeli submarines. Precise energy density of these batteries is not publically available.  For domestic uses such as handheld electronics or solar power applications, lithium ion batteries may have 3-5 times the energy density of lead-acid batteries \parencite{oconnor2017}.  However, the actual useful density of lithium ion batteries can vary widely depending on different chemistries and designs.  It might be assumed that for very large battery banks (such as submarines) where the stability of the battery bank is of critical importance, maximum density migth be sacrificed in favour of additional safety systems, such as increased monitoring, as well as spacing and insulation barriers between cells, in order to prevent thermal runaway in one cell leading to a catastrophic fire.  The realistic energy density advantage that might be obtained as been suggested to lying closer to a factor of two or slightly less \parencite{greenfield2016}.

@@

The phenomenon manifests itself in a very substantial reduction in the overall range that can be achieved submerged, and a difference between the speed at which maximum range can be achieved.  Under central assumptions, the propeller driven submarine can reach over 540nm at any speed between 4 and 5kt. The pump-jet driven submarine reaches over 360nm between 3 and 6kt. A difference of 180nm dived range would presumeably be of considerable tactical significance.

@@

The most striking feature of this figure however is the impact on the submarine's performance when hotel load is lowered, and battery density is raised simultaneously.  Here the two factors combined have a proportional impact of around 80\% improvement for the switch to propellers under central assumptions, and over 150\% in the scenario of widest separation.  However, in nominal terms, this impact is even more significant, with central scenarios showing a \textit{difference} of over 200 hours between the two systesm under central assumptions at 3kt, and a range difference of over 750nm.  These differences are extremely large, representing plausibly double the expected total ranges and endurances of existing modern diesel-electric submarines. Even if such large changes in hotel load and battery capacity are not simultaneously likely these sensitivitiesquite clearly demonstrate how even modest shifts in both hotel load and battery energy density might combine to have very significant impacts on the appearnce of the pump-jet and propeller comparison.  For instance, a hotel load that reduced by 30\%, and a battery energy density increase of 40\% could plausibly have the impact of doubling the nominal differces between the performance of the two propulsive systems.

@@

The magnitude of the difference is also substantial.  For the central efficiency curve and hotel load assumptions, the pumpjet would cause a sumbarine to increase the time spent snorting for by over 60\% between 5kt and 7.5kt.  One would expect that this would be of considerable tactical consequence. These graphs also demonstrate that the difference does not diminish at higher speeds, infact it begins to increase again above 10kt, as the power required for propulsion begins consume much of the power generated by the diesel engines, and efficiency difference strongly impacts the size of the diminishing surplus available for charging.  It is also clear that the sensitivity to the selection of efficiency curves remains quite strong in this case.  It is clearly plausible that in a range of circumstances the pumpjet could double the time spent snorting.  However, the difference between the lines representing the pair of central efficiency curves, and the narrowest separation, is noticeably smaller, and the impact of chaning hotel load on these lines is also proportionally diminished.  As such, it might be said with considerable confidence that the most likely impact of the pumpjet on indiscetion ratios at any speeds over 4kt will around 40\% or higher.  It would be extremely unlikely that the penalty induced by the pumpjet at speeds of 5kt and above would be any less than 25\%.

>>>>

It is clear that the optimisation of a system for minimising cavitation, and hence improving acoustic performance, is a separate and independent design requirement to that of optimising a system for propulsive efficiency.  These two design requirements can be found either to be in tension, or alignment, in different regimes.  At high speeds, where cavitation tends to have severe impacts on propulsive efficiency of open propellers, and high mass-flow (which leads to good efficiency) can be achieved through compact ducts and pumps, pumpjets offer very good performance on both metrics.  Hence, in this regime the two design requirements tend to align relatively well in favour of pumpjets.  However, in very low-speed, low-load environments, where cavitation effects are negligible, and larger areas need to be engaged achieve good mass-flow and efficiency, the two design requirements align relatively well in favour of an open propeller.  In a variety of regimes (including at quite high speeds), the two requirements find themselves in tension and needing to be balanced \parencite{gearhart1966selection}.

@@

For context, the equations developed by Wislicenus are used for the purposes generally of selecting the ideal propulor diameter, which can be made larger in the case of a propeller to access higher ideal efficiencies.  This is an opportunity which is constrained in the case of jets because the efficiency diminishes at very low values of $\Delta V / V_0$.  This is quite a different constraint to other "practical" considerations as some have suggested, such as weight, or space constraints.  Open propellers may advantageously choose to lower their values of $\Delta V / V_0$ as low as 0.1 in order to obtain ideal efficiencies \parencite{wislicenus1973}.  This still provides a hard limit on potential improvements that can be attained for a pumpjet that has non-zero duct loss.

@@

In order to begin to model the likely impact of propulsion system choices on system peformance characteristics such as dived range and endurance, it is necessary to adopt some plausible curves indicating likely levels of efficiency across the operating speed range, in this case assumed to be up to 20kt.In this case, three possible curves have been selected for each system, representing a high, low, and central assumption for each, with peak efficiency for the propeller at about 65\% at near 5kt, and jet efficiency around 27\% at the sames speed.

The central assumption has been taken in both cases from the study undertaken by BMT and Rolls Royce, as in \cite{giles2010}, and indicated in \ref{fig:BMTShape.png}, since this represents the most direct comparison made as a simple function of waterspeed, with comparable technologies. It is probable that the pumpjet of the submarine will be a purely axial-flow pump, as opposed to a mixed-flow pump, but this difference in turbomachinery is unlikely to result in a substantial loss of propulsor efficiency, as discussed in Section \ref{pumptypes}. Having the outflow of the waterjet released entirely underneath the surface (as opposed to just above it, in the case of most waterjets) certainly removes the most substantive difference that might drive significant changes in the flow-rates, diameters, and other significant determinants of the machinery characteristics that might normally distinguish different types of pumps.  In addition, the intake duct may be somewhat longer, increasing duct losses, but the degree of diffusion in the duct may be lower, reducing the adverse pressure gradient.  In other words, it is plausible that different individual components may incur quite different levels of certain types of losses, but it is unlikely that the overall efficiency profile of the system as a whole is dramatically altered assuming competent design, since an enclosed intake controlling the waterspeed at the impeller, and contracting nozzle beyond will remain common features of both. Indeed, it is difficult to imagine why the US Navy, having already attempted a pure axial-flow pumpjet directly in place of propellers for similar ship for the same purpose (reduction of cavitation-induced noise for Anti-Submarine Warfare) in 1965 \parencite{harvie1965construction}, would support a study 40 years later using a pump type that was expected to have substantially reduced efficiency at the intended operating speeds.

Small design difference between the propellers used on submarines and surface ships may also exsit. It is probable that the the ship propeller would have five blades, where the submarine would likely have seven or nine. However, being for a warship designed to travel at similar speeds to those that a nuclear submarine might, with relatively unconstrained propeller size, the impact of the propeller difference here are is still likely to be very small relative to the contrast with jets.

Overall, however, its considered that having an experimentally verified curve comparing very similar technologies as a direct function of waterspeed, on vessels attempting to optimise for the same characteristic (reduced noise induced by cavitation) provides a far more certain basis for comparison than attempting to derive efficiency curves \textit{ab initio} using theortical dimensionless factors corresponding and our best estimate of a more likely precise design.

Variations from this central line have been made with consideration for some plausible trade-offs errors, which are intended to be indicative of magnitude possible unknowns in system design and their likely impact.  In the case of the 'high' pumpjet assumption, it is assume that a substantially larger intake area has been chosen in order to achieve higher efficiencies at lower speeds, at the cost of peak ideal efficiency, which might be realised not too far above 20kt for such a design.  The low assumption for the pumpjet is still represents higher efficiencies at equivalent speeds than those given in \cite{fujisawa1995} for a waterjet. It is assumed that for both the central and low pumpjet curves, peak efficiency would not be achieved until a speed considerably above 20kt.  All of the efficiency curves reflect a smaller decline in efficiency than the indicative assumption used by Andrew Davies, namely that efficiency might reduce by a factor of four between transit and patrol speeds, if those speeds also differed by a factor of four \cite{davies2017}.

@@

It is worthwhile considering that alternative types of battery may be on offer for submarines.  In particular, lithium ion batteries have been selected by Norway in a selection process which happend almost concurrently with Australia's \parencite{ohff201702}, and would probably have been on offer to Australia if they had been sought. Precise energy density of these batteries is not publically available.  For domestic uses such as handheld electronics or solar power applications, lithium ion batteries may have 3-5 times the energy density of lead-acid batteries \parencite{oconnor2017}.  However, the actual useful density of lithium ion batteries can vary widely depending on different chemistries and designs.  It might be assumed that for very large battery banks (such as submarines) where the stability of the battery bank is of critical importance, maximum density migth be sacrificed in favour of additional safety systems, such as increased monitoring, as well as spacing and insulation barriers between cells, in order to prevent thermal runaway in one cell leading to a catastrophic fire.  The realistic energy density advantage that might be obtained as been suggested to lying closer to a factor of two or slightly less \parencite{greenfield2016}.

@@

The phenomenon manifests itself in a very substantial reduction in the overall range that can be achieved submerged, and a difference between the speed at which maximum range can be achieved.  Under central assumptions, the propeller driven submarine can reach over 540nm at any speed between 4 and 5kt. The pumpjet driven submarine reaches over 360nm between 3 and 6kt. A difference of 180nm dived range would presumeably be of considerable tactical significance.

@@

The most striking feature of this figure however is the impact on the submarine's performance when hotel load is lowered, and battery density is raised simultaneously.  Here the two factors combined have a proportional impact of around 80\% improvement for the switch to propellers under central assumptions, and over 150\% in the scenario of widest separation.  However, in nominal terms, this impact is even more significant, with central scenarios showing a \textit{difference} of over 200 hours between the two systesm under central assumptions at 3kt, and a range difference of over 750nm.  These differences are extremely large, representing plausibly double the expected total ranges and endurances of existing modern diesel-electric submarines. Even if such large changes in hotel load and battery capacity are not simultaneously likely these sensitivitiesquite clearly demonstrate how even modest shifts in both hotel load and battery energy density might combine to have very significant impacts on the appearnce of the pumpjet and propeller comparison.  For instance, a hotel load that reduced by 30\%, and a battery energy density increase of 40\% could plausibly have the impact of doubling the nominal differces between the performance of the two propulsive systems.

@@

The magnitude of the difference is also substantial.  For the central efficiency curve and hotel load assumptions, the pumpjet would cause a sumbarine to increase the time spent snorting for by over 60\% between 5kt and 7.5kt.  One would expect that this would be of considerable tactical consequence. These graphs also demonstrate that the difference does not diminish at higher speeds, infact it begins to increase again above 10kt, as the power required for propulsion begins to consume much of the power generated by the diesel engines, and the efficiency difference strongly impacts the size of the diminishing surplus available for charging.  It is also clear that the sensitivity to the selection of efficiency curves remains quite strong in this case.  It is clearly plausible that in a range of circumstances the pumpjet could double the time spent snorting.  However, the difference between the lines representing the pair of central efficiency curves, and the narrowest separation, is noticeably smaller, and the impact of chaning hotel load on these lines is also proportionally diminished.  As such, it might be said with considerable confidence that the most likely impact of the pumpjet on indiscetion ratios at any speeds over 4kt will around 40\% or higher.  It would be extremely unlikely that the penalty induced by the pumpjet at speeds of 5kt and above would be any less than 25\%.
